---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(rRMSAnalyzer)
```

# rRMSAnalyzer: package to analyze RiboMethSeq data

RiboMethSeq is an RNAseq-based approach to analyze 2'O-ribose methylation (2'Ome).

rRMSAnalyzer is an R package that provides a set of easy-to-use functions to evaluate 2'Ome levels by computing C-scores from RiboMethSeq read end counts as input.

Available features (version 2): 

- C-score computation (using either mean or median for the window of neighboring positions)
- Batch effect adjustment with CombatSeq
- Different visualizations to compare samples or sites
- Include a table of annotated human rRNA sites 
- Export computed C-scores into a dataframe 
- Semi-automated quality control report

> **Note** We have also developed a [dedicated Nextflow pipeline](https://github.com/RibosomeCRCL/ribomethseq-nf) to process the data from sequencing output (fastq files) to useful raw data for rRMSAnalyzer (read end counts).

## Installation

The latest version of rRMSAnalyzer package can be installed from Github with:

```{r, eval=FALSE}
library(devtools)
devtools::install_github("RibosomeCRCL/rRMSAnalyzer")
```

## Usage

```{r, eval = FALSE}
library(rRMSAnalyzer)

ribo <- load_ribodata(
              count_path = "/path/to/your/csvfiles/directory/",
              metadata = "path/to/metadata.csv",
              metadata_key = "filename",
              metadata_id = "samplename")

# Compute the c-score using different parameters,
# including calculation of the local coverage using the mean instead of the median
ribo <- compute_cscore(ribo, method = "mean")

# If necessary, adjust any technical biases using ComBat-Seq.
# Here, as an example, we use the "library" column in metadata.
ribo <- adjust_bias(ribo,"library")

# Plot a Principal Component Analysis (PCA) whose colors depend on the "condition" column in metadata
plot_pca(ribo,"condition")
```

## Getting started

The "getting started" is available on our website: https://ribosomecrcl.github.io/rRMSAnalyzer/

A test dataset (ribo_toy) is included in the package.

## Help, bug reports and suggestions

To report a bug or any suggestion to improve the package, please let us known by opening a new issue on: <https://github.com/RibosomeCRCL/rRMSAnalyzer/issues>

## Acknowledgements

We would like to thank all our collaborators from Jean-Jacques Diaz Team and the Bioinformatic Platform Gilles Thomas for their advices and suggestions.

## Funding

This project has been funded by the French Cancer Institute (INCa, PLBIO 2019-138 MARACAS), the SIRIC Program (INCa-DGOS-Inserm_12563 LyRICAN), LabEX program (DEVweCan), the French association Ligue Nationale Contre le Cancer and Synergie Lyon Cancer Foundation.